How to publish in Biological Sciences: a guide for the uninitiated

### 2021-06-10 1.08-advisorless.Rmd
Add files via upload
# UPLOAD working version that prints to pdf and gitbook

# Advisor-less {#advisorless}

During your PhD you will have had an advisor who would have been central to many of the decisions that you made. However your relationship with this person ended up, it is likely that you will still need advice from someone more experienced. If you are a postdoc, you should still have an academic whose research group you are working in, but as an early career researcher you may feel that you're suddenly cut adrift without advice. Increasing numbers of institutions are now recognising that early career researchers need mentors to advise them about how to move forwards within their academic field.

@@

Doesn't have to be in the same institution, or even in the same country. In these days of post-COVID communication we are all a lot happier hooking up to meetings online.

@@

Where to publish / What meeting to go to / networks

>>>>

# Advisor-less in the academic world - growing your network {#advisorless}

During your PhD you will have had an advisor who would have been central to many of the decisions that you made. Don't give up on that relationship, it is something that you have at the heart of your network. Therefore it is worth maintaining in the best condition possible as this is someone that you will need to go back to (probably lots of times) for references and other professional help. In short, always keep your advisor sweet.

Having said this, as an early career researcher, you are now advisor-less, yet it is likely that you will still need advice from someone more experienced. Hopefully, there is always your advisor, but the chance is that you may well have changed institutions, and so they may not be so close or easy to contact. This is a good reason for expanding your network by looking for a mentor to your role as an early career researcher. Increasing numbers of institutions are now recognising that early career researchers need mentors to advise them about how to move forwards within their academic field.

@@

The first thing to look for in a mentor is someone that you feel you are able to talk to easily. As you will see, there are a wide range of topics to discuss in terms of building your career, and you will need to find someone that you feel you can talk to about nearly everything safely and without fear that it might be used against you. The mentor should feel happy to talk and spend time with you. They should be someone that is genuine in their desire to help you and your career. You won't have a fruitful relationship with someone that doesn't really like you, or doesn't want to spend time talking to you.

Next, your mentor should be someone who is already well established within an academic network that you want to (or already are) a member of. You will need to have common ground to talk about, and getting their advice on building your network will necessitate talking about how to manage others in the network.

The personal touch is nice to have, so if you can find a mentor who has an office in the same building, or on the same campus, then it's great to be able to meet up and chat. However, your mentor doesn't have to be in the same institution, or even in the same country. In these days of post-COVID communication we are all a lot happier hooking up to meetings online.

If you still have choices, my suggestion would be to look for someone that you share non-academic interests with. For example, you may both enjoy playing squash together, attending opera or scuba-diving. Meeting outside academia will make your mentor-mentee chats easier for both of you. Otherwise, try not to make your meetings with your mentor a burden on their work time. All academics are busy and core-working hours are hard to come by. Instead you could suggest taking your mentor out for a coffee or a sandwich in which to have your chat.

Note that a mentor doesn't have to be someone that is particularly senior, although they should be well connected. Having said this, they are likely to be at a more advanced career stage than you.

@@

Your mentor's time is valuable, so always have an agenda when you go to talk to them. Have ideas that you want to pitch and get feedback from. Have some way of making notes about what they say, especially if you meet over a beer. You don't want to have to go back and ask them again for the same advice. Be prepared to get the answer that they give. This seems an odd thing to say, but they may not always like your idea, and you should be prepared to listen to their advice whatever it is, not only when it agrees with your own ideas.

Typical things to discuss with your mentor are:

- Where to publish
- What meeting to attend
- How to extend your network
- Whether (or not) to write a reply or commentary on another groups' work
- Whether to respond to a call for grant applications
- Whether to apply for another job

You may also want to share some of the more exciting aspects of your work, your findings and how you interpret them. Try to keep away from moaning about other academics, the amount of administration you have, teaching burdens, etc. Make your discussions as positive as possible so that your mentor is more likely to want to maintain and grow your relationship.

## The importance of networks in your academic career {#networks}
There is clear evidence that researchers with good networks collaborate more, publish more and are better cited [@parish2018dynamics]. Your mentor can be one way to increase your academic network. Perhaps more importantly, they can be a guide to help you through the network that you are both involved with.

There is a good chance that you are already in a network where your advisor, and perhaps some members of your thesis committee sit at the hub of an extended network that you have had partial exposure to during the course of your studies. If you haven't realised already, these networks are of profound importance in every step of your academic career.

**Benefits of a good network include: **
 - increased citations for your work
 - better chances that your work will be edited or reviewed by a member or an associate of your network, resulting in a higher chance of publication or a grant application accepted
 - increased potential to be nominated for an award
 - increased potential to be invited to give talks at conferences
 - invitation to apply for positions and jobs in the labs of partners at institutions in the network

If you are already in a network, you should be thinking of ways in which you can get more from your network now that you are an early career researcher. If, on the other hand, you are not in a good (sizeable and supportive) network, you should be thinking about how to join an existing network or how to form a network around yourself and your work.

In many cases, managing your network is about providing opportunities for yourself and others in your network. There are many ways in which you can do this, but you should be prepared to put in a lot of work from your own side.

**Potential initiatives include:**

- writing a grant application that involves others in your network
- leading a piece of written work that includes others in the network
  + a review
  + a commentary on another article (not from your network!)
  + a viewpoint or position article on an emerging topic
- organising a symposium aimed a central or emerging question within your network
- organising a special issue of a journal (potentially as a result of the symposium)

Your chosen mentor is the best person to discuss how to make waves in your network. Remember that relationships have already been forged in networks and you should never try to destabilise these. Instead, you should aim to become another node of the network by growing your own usefulness as a positive force for new and interesting initiatives that others will want to join. Because you are new, don't expect to be included in every initiative. If you feel that you are being left out of other initiatives that you've heard about unnecessarily, ask your mentor whether or not it is appropriate for you to bring this up.

Always bridge and build relationships. As you should have become aware by now, there are many potential ways of becoming involved in combative elements of academic life. Your own sub-discipline will likely have groups of people that write commentaries about the work of other groups. Networks are powerful places to be, but they can also become destructive when another network opposes a position. I have seen this type of behaviour happen within and around my own limited subdisciplines. One group will publish a major finding in _Nature_ or _Science_ and a week or two later, the other group will publish a commentary pointing out what are (to them) flaws in the first group's publication. I always have a sneaking admiration for those individuals that appear to swim between these groups, always maintaining good relationships with both sides. Theirs is really an interesting, but perhaps precarious, position.

If you feel that you don't have any network, then I think that you should reassess your position. You likely do have people that you talk to and interact with that are already part of a network, but your question is how to increase the size, and potentially scope, of your network. One point that you will will see in [Part IV](jobinacademia.html) is that everyone will have their own path through their career. There really is no one size fits all way. Growing your network is similar in that each academic will do this in ways that are unique to them.

### Calculating the size of your network
A simple metric for your network is the number of co-authors that you have. Clearly, as you increase the number of people that you work (and hopefully) publish with, you will have a bigger, better functioning network. However, scientometricians have their own ways of interpreting and mapping networks (see Figure \@ref(fig:Olaf-SA-invasive-fish)).

(ref:Olaf-SA-invasive-fish) **An author network for the Web of Science search for "Invasive fish" AND "South Africa"**. In this author network you will see my friend and colleague, the late Prof. Olaf Weyl, who is at the centre of a large group of people who worked on invasive fish in South Africa. Olaf's wasn't the only research group, but certainly the most influential and joined up. Olaf's collaborations were extensive, spanning continents and generations of researchers. Drawn with VOSviewer [@vaneck2010software].

```{r Olaf-SA-invasive-fish, echo=FALSE, out.width = '85%', fig.cap="(ref:Olaf-SA-invasive-fish)"}
 knitr::include_graphics("figures/Olaf-SA-invasive-fish.png")
```

If you aren't sure about who's in what networks within your own researcher area, you can use appropriate keywords to download literature within the area that you work, and plot (using [VOSviewer](https://www.vosviewer.com/)) a similar author and co-author network. I'd suggest using enough keywords to call ~1000 articles (including their citations) that include your research area (all or the majority of your papers should be included). Using a _map based on bibliographic data_, read the bibliometric files and analyse using co-authorship to describe the linkages between authors. This can be very revealing (see Figure \@ref(fig:Olaf-SA-invasive-fish)), and while you should already be well aware of the big names in your area, some of the people that span areas and sub-subdisciplines might well be worth seeking out. When looking into their publications, ask yourself how they managed to span between networks (conducting a postdoc in each lab is one such way). More than this, you can use collaborations to combine networks of other Early Career Researchers in your subject area.

Another method for determining the size of your network, or those of potential collaborators around you, is to calculate your R value [@ioannidis2008measuring]. **R** is made up of two figures, _I~1~_ which is the number of authors that appear in at least _I~1~_ papers, divided by _N~p~_ which is the number of papers that the author has published. _I~1~_ is a little like the H-index in that it increases each time you a person in your network. For example, to have a _I~1~_ of 4, you need at least 4 publications in which you and the same 4 other authors occur. Parish et al [-@parish2018dynamics] show that as _I~1~_ grows, so the value of **R** decreases and researchers become most productive as **R** approaches 1.

Once you start to calculate your own value of **R**, you will appreciate that not only are large networks important, but that it is useful to start these networks early and maintain working with an every increasing group throughout your career.

### Using social-networks to grow your nework
I think that the ways in which you can grow your network have never been so great. Social networking is a massive way to reach other researchers and you don't have to wait for a meeting to do this. The #academictwitter groups are already very large and established. Investing time in getting connected is worthwhile, especially when you can meet up in person through conferences and symposia. I don't want to make out that Twitter, and (no doubt) other social networking is easy, but it is a good way of hooking up with a whole lot of other people without having to leave your desk. You can pitch ideas, and get collaborations moving in your area. Once you have it working for you, it is very empowering. If social networking is your thing, then I'd suggest trying to find yourself a 'social networking mentor' - someone that already has a good following and can guide you through the various nuances of rules that happen on these platforms.

### 2021-06-10 1.09-preprints.Rmd
Add files via upload
# UPLOAD working version that prints to pdf and gitbook

So far in this book I have explained about the process of peer review how important this is and why it's the gold standard in science. Preprints came to us from the world of physics. An academic world that is moving so quickly that many inside it don't want to wait for peer review before making their work public.

Preprints do a little more than this because it not only makes the work public but allows it to be read by more peers. Many consider it as a kind of open peer-review system.

@@

@@

@@

@@

As far back as 2002, William Arms [@arms2002what] suggested that openly soliciting comments on the web might be an alternative for peer review of scholarly articles. Sixteen years later, this has come to pass in the form of preprints.

@@

No. Reviewers are chosen with great care because the area is in their particular domain. They have insights that not everyone will be aware of and these are an important aspect of the purpose of peer review.

I edit for the journal PeerJ. Although there can be various reasons to be rejected from PeerJ, normally it means that your paper is not scientifically sound. As PeerJ has no selection for impact, rejection does not normally mean that it can be simply submitted to another journal. I have noticed that manuscripts that I have rejected from PeerJ are still available as PeerJ-preprints without any comment on their failure to go through peer review. In my opinion, this is not good as it essentially ignores the input given by both reviewers and editors. The article appears as if it has had no comments or attention, when this is not the case. In a system where we move to relying more on preprints, why would we want to ignore chosen peer reviewers for whom this article was within their specialist area?

@@

## In my opinion, comments pages won’t replace peer review.

If we end up abandoning our current way of publishing in favour of a comments page, I think that we’ll all be worse off. Much more likely is that increasing numbers of journals will require preprints as well as open peer review, hopefully leaning towards a more open and transparent publishing experience.

We simply need to overcome the vanity of having our manuscripts set out in a pretty way. If you are a postgraduate student reading this, then you are the first who will suffer from the mistakes of your community over the past 40 years. But you can also be the solution, mostly because you will be so badly in need of a solution that the system has to change for academia to survive. Us oldies will eventually be happy to accept unformatted manuscripts as the way science is presented, I think we’ll be able to move on without the involvement and the exorbitant costs that the publishers extort from us. What will be a bigger challenge is to topple the provaledge from those who control STEM with the editorial roles. The temptation will always be for the younger elite to take on that privilege, to benefit as previous generations of academics have. Breaking that circle of abuse will require widespread adoption of new rules for peer review.

>>>>

So far in this book I have explained about the process of peer review how important this is and why it's the gold standard in science. Preprints came to us from the world of physics. An academic world that is moving so quickly that many inside it don't want to wait for peer review before making their work public. They date back to 1991, and were the brainchild of Paul Ginsparg with his preprint server, arXiv. Today arXiv hosts nearly 2 million articles in 8 subject areas: Physics, Mathematics, Computer Science, Quantitative Biology, Quantitative Finance, Statistics, Electrical Engineering and Systems Science and Economics.

In addition to making your work open access, it allows anyone to read and review it. This is unlike the traditional publishing model where editors invite selected reviewers. Many consider preprints as a kind of open peer-review system.

> "The life science community needs to return to a culture of evaluating scientific merit from reading manuscripts, rather than basing judgment on where papers are published."
>
> ---@vale2015accelerating

@@

## Who posts preprints on bioRxiv? {#bioRxiv}

In an analysis by Abdill and Blekhman [-@abdill2019tracking] back in 2018, there were 37648 preprints uploaded to bioRxiv. But in the world of preprints, this information is quite dated as back then there were only ~2000 uploads each month. Luckily, these same authors regularly scrape data regarding submissions to bioRxiv and I'm able to reproduce these live data below. It is interesting to look back at the concern back in 2015 that preprint servers like bioRxiv might not catch on in the biological sciences [@vale2015accelerating], and it is still true to say that the number of preprints in the life sciences is still dwarfed by the annual number of publications, whereas physics has seen the opposite trend.

(ref:bioRxiv-line-annotated) **The growth of bioRxiv submissions in time.** Note how the rate of submission rocketed at the beginning of the pandemic lockdown, and has remained high ever since.

```{r bioRxiv-line-annotated, echo=FALSE, fig.cap="(ref:bioRxiv-line-annotated)"}
if(knitr::is_html_output(excludes="markdown")) knitr::include_url("https://johnmeasey.github.io/bioRxiv-line-annotated/index.html", height="350px") else knitr::include_graphics("figures/bioRxiv-line-annotated/annotated-line-example.png")
```

What you see is a doubling of bioRxiv submissions dating back to May 2020 (Figure \@ref(fig:bioRxiv-line-annotated)), a few months after the start of the global COVID pandemic. (Go to the [RXivist website](https://rxivist.org/stats) to see the incredible spike in [medRxiv](https://www.medrxiv.org/) data at the same point in time.) Many scientists had spent a few months working from home. Some had been productive, and many decided to move this productivity for the first time onto bioRxiv.

In answer to the question above, **just about everyone now posts preprints on bioRxiv**. Those that don't likely use other preprint servers, or are not moving with the times.

The preprint revolution has not gone unnoticed by the tech giants. Back in 2017 the [Chan Zuckerberg Initiative](https://chanzuckerberg.com/) made an undisclosed donation to bioRxiv. The advisory board also has the architect for Google Scholar, Anurag Acharya.

@@

Posting a preprint on bioRxiv is also a shortcut to submission to a growing number of traditional journals. Manuscripts and Supplementary Information can be [transmitted directly](https://www.biorxiv.org/about-biorxiv) from the preprint server to many journals without the need to upload the files and metadata a second time.

@@

If you choose to publish in an [overlay journal](#overlay), then you'll have to deposit your submission onto a preprint server.

@@

As far back as 2002, William Arms [-@arms2002what] suggested that openly soliciting comments on the web might be an alternative for peer review of scholarly articles. Sixteen years later, this has come to pass in the form of preprints.

@@

No. Reviewers are chosen by an editor with great care because the area is in their particular domain. They have insights that not everyone will be aware of and these are an important aspect of the purpose of peer review.

I edit for the journal [PeerJ](www.peerj.com). Although there can be various reasons to be rejected from PeerJ, normally it means that your paper is not scientifically sound. As PeerJ has no selection for impact, rejection does not normally mean that it can be simply submitted to another journal. I have noticed that manuscripts that I have rejected from PeerJ are still available as PeerJ-preprints without any comment on their failure to go through peer review. In my opinion, this is not good as it essentially ignores the input given by both reviewers and editors. The article appears as if it has had no comments or attention, when this is not the case. In a system where we move to relying more on preprints, why would we want to ignore chosen peer reviewers for whom this article was within their specialist area?

@@

## Preprints are here to stay

It is clear that preprints are with us to stay. The year of the COVID-19 pandemic (2020) saw an explosion of preprint papers on the topic, but also saw the misunderstanding of what these articles mean by the press and general public alike. Rapid sharing of results via preprint servers has already been put in place following the outbreak of Zika virus in South America (back in 2015-16), but the global nature of the Covid public health crisis saw much larger numbers of preprints being placed online.

But the value of preprints will always be limited for as long as there is no peer review. Moreover, comments won't suffice for peer review as there is no editorial oversight. As you'll read elsewhere, the [role of the editor](editors.html) is pivotal in publishing.

## The exciting new world of Overlay Journals {#overlay}

Having said that preprints won't replace the role of peer review, what if we did have good, editorially coordinated peer review of preprints. What if, instead of these manuscripts effectively leaving the preprint system, they were updated together with the reviews that prompted the updates, each with their own linked DOI. What if the journals themselves were simply pointing to collections of papers that had been curated in this way? Simply a website that throws a veneer of a journal as waypoints to peer reviewed journals? This world has already been imagined and is functioning in mathematics, where Overlay Journals have begun to prosper.

According to Brown [-@brown2010introduction], the idea of overlaying has been with us for some time, and exist as websites that offer a series of links to other papers. In this way, a review article could be considered an 'overlay paper', the contents of Web of Science as an 'overlay database'. But, for me at least, this is not where the real potential lies. Instead, imagine the overlay journal as a way in which academics entirely remove the need for publishers. The need for this is increasingly evident as we become more familiar with the ways in which we rely on traditional publishing models to pervade our scientific project with confirmation bias. Overlay journals no longer require a publisher to store the publication. This is done at the preprint server. The reviews are housed at the same arXiv site (or would be in an ideal and transparent version)[@rittman2020fast], as is the manuscript in its final form after being accepted by the overlay journal editor. The authors themselves are responsible for the final layout. The Overlay Journal co-ordinates the reviews and conducts the editorial work, and then simply acts as a pointer to the finished product: no papers, no publishers, no editorial management software, no costs and all papers are [Diamond OA](#diamond)!

The journal [_Discrete Analysis_](https://discreteanalysisjournal.com/) (indexed in both Web of Science and Scopus) was the first of these new 'arXiv overlay journals' (since 2017), and following on this link will allow you to quickly appreciate what an Overlay Journal is. Each 'published' paper still sits on its original preprint server. The overlay journal itself offers a brief editorial summary of what you'll find if you click through to the paper. This is a fantastic idea in that it pitches editors back into being responsible content curators. As an editor I'd want to be motivated to publish a paper that I liked in order to write an editorial summary about it.

Because the only the accepted version is provided with an 'article number' and the style file of the journal layout, the author then produces the final version of record (VoR) of the accepted manuscript by running the style file with LaTex. All of this is possible with free software, for example by using Rmarkdown [@xie2018r].

Using preprint servers also allows the entire process to be [transparent](#transparency2), very quickly becoming associated with other great initiatives like the Centre for Open Science - [OSF](https://www.cos.io/).

### What do traditional publishers think of 'Overlay Journals'?

Surely, the onset of 'Overlay Journals' should have publishers quaking in their boots? Strangley not. But their response should really be enough to wake us up.

> They're probably only going to succeed in disciplines where a no-frills approach to publishing is acceptable. 
>
> [Anonymous from Highwire Press Inc. 2017](https://youtu.be/ADecAXLrKoA)

But they do see that there's a possibility of disruption:

> I think that the real threat to our traditional model... if Overlay Journals have **Impact Factors** and can provide the same services, and they are free... then I think that that does pose a threat.

As this has already happened, it would be interesting to know how traditional publishers are going to prevent an Overlay Journal take-over.

### What is happening in biological sciences? {#PCI-EvolBiol}

One of the original electronic journals, [Journal of Medical Internet Research (JMIR)](https://www.jmir.org/), announced in 2019 that it will launch an overlay journal covering biology, a so called 'superjournal', JMIRx [@eysenbach2019celebrating]. This overlay journal operates by editors choosing preprints that they want to publish ('editorial prospecting'), and then approaching authors and reviewers, and also by authors pitching their preprints to the editors. Today [JMIRx|Bio](https://bio.jmirx.org/) accepts any preprint published on [bioRxiv](#bioRxiv). Although JMRIx|Bio, and sister journal [JMRIx|Psy](https://psy.jmirx.org/), were launched in 2020, I cannot find any articles submitted (by mid-2021). The sister journal [JMIRx|Med](https://med.jmirx.org/), launched at the same time and in the same area as other JMIR journals, has a rapidly expanding publication base.

There's an excellent tie-in here with transparency. Because preprints are Diamond OA, and reviews are OA, the process is all transparent.

A nearly 'overlay' model is the [Peer Community in Evolutionary Biology](https://evolbiol.peercommunityin.org/), launched in Janurary 2017. This comes very close to the 'arXiv Overlay Journal' model described above. These preprints are submitted to PCI-Evol Biol, and are reviewed and (if they aren't rejected), a recommendation is given. The site then publishes the recommendation from peers as well as pointing to the preprint. However, unlike _Discrete Analysis_, the preprint remains 'unpublished' despite the peer review and can then be taken onto a traditional journal. There is a  growing [list of journals](https://evolbiol.peercommunityin.org/about/about) whose editors will accept recommendations from PCI-EvolBiol, and may use the reviews when appropriate. However, it's also worth noting that there are a small number of journals that will not accept preprints recommended by PCI-EvolBiol. While [Peer Community in Evolutionary Biology](https://evolbiol.peercommunityin.org/) does not publish their peer reviewed articles, another initiative from [Peer Community In](https://peercommunityin.org/) takes a step backwards to get a step closer.

The journal [eLife](https://elifesciences.org/) is also taking steps toward becoming an Overlay Journal, with their implementation of a **preprint only** submission route [@eisen2020implementing]. Although eLife appears to embrace all the advantages of transparency in their use of preprints, there is still a significant barrier that has recently jumped from $2500 to $3000 for the privilege to publish (eLife has an APC waiver system which is not seen by editors). Again the question of what exactly scholars are paying such high fees for comes to the forefront.

### Preregistration and a commitment to publish {#commitment}
Registering your proposal (or any research plan), means that you can present a historical document to a journal (probably 4 to 5 years later) to show that you have tested the hypotheses that you originally intended to. This is simply a way of being transparent in your science, and enables you to demonstrate that you have not been [p-hacking or HARKing](transparency.html). Similarly, you can demonstrate that you are not '[salami-slicing](#salami)' your results.

Another initiative from [Peer Community In](https://peercommunityin.org/) is the possibility of submitting to their [Registered Reports](https://rr.peercommunityin.org/), which goes much further towards removing the confirmation bias. The Registered Report (RR) is in effect the registration of a proposal (i.e. preregistration) with review. If the RR is approved by reviewers then the study is, in principle, given the green light for publication whether or not the hypothesis posed is accepted or rejected. I say “in principle” because those same reviewers are shown the ms again once the results are in. They need to check that the methods proposed were followed, the analyses were conducted in the same way they were proposed, and that the conclusions are justified by the results. In terms of figure 27.1, Peer Community In are offering to organise the two yellow review arrows. In addition, there are a bunch of journals that have already signed up to accept RRs that are signed off after completion (notable among these is PeerJ). To me, this represents an important step in the right direction toward [transparency](transparency2.html), and the elimination of confirmation bias. What would be great to see is the number of conventional journals sign up with RRs based on the quality of the study design and execution, and the concomitant abandonment of [Impact Factor](impactfactor.html) as a driving force in publishing.

### Beware of publishers' preprint servers
Publishers have a lot to be worried about preprint servers as they have the potential to take away their business. For this reason, you will see that some publishers have launched their own preprint services. Don't be lulled into submitting to the preprint server of a publisher. There are plenty of preprint servers that have nothing to do with publishers, including the most prestigious. These are non-profit transparent organisations, and we all have an interest in them staying that way. Simply put, there's no need to use a publishers' preprint server (this is the same principle as your data, [see here](#suppmat)).

## Peeriodicals - another twist on the idea of an Overlay Journal

The launch in 2018 of [Peeriodicals](https://peeriodicals.com/) puts another twist on the idea of an Overlay Journal. This time, the Peeriodical is a site where anyone puts together a collection of any published papers or pre-prints that they curate themselves on a topic. They don't pass them out for any further review, but they are in effect another kind of curated Overlay.

### 2021-06-10 1.10-review.Rmd
Add files via upload
# UPLOAD working version that prints to pdf and gitbook

The first chapter of many PhDs is often a place for a literature review. Writing a review of the literature is actually a very useful task to undertake during your PhD, preferably towards the beginning. Firstly it will get you familiar with all of the relevant literature in your field. It will give you some idea of the historical context of your subject area. And it's especially good to help you spot gaps in the literature that may not have been realised before.

@@

## Having a purpose
Writing a literature review is easiest when you have a clear purpose. The objective can be particularly tight, or can aim to use the literature in order to answer a broader question. Either way having a clear purpose for your literature review will help you to define exactly what literature you're looking for.

Without an aim a literature review can be extremely daunting. There will be literally no end to the literature that you could include and it may well overwhelm you. If your question is more broad it will help to define a clear time period in which you are searching for literature. For example you may want to just consider the last 10 years. Remember that the amount of literature is increasing almost exponentially and so if you are covering the same ground as a previous review a lot can happen in 10 years.

@@

## Systematic reviews
I prefer to conduct a systematic review. This is where you define your search criteria and the databases that you want to use and then include all of the literature that comes out. You will need to provide clear criteria for the inclusion of literature. As well as explaining exactly how you decide to include or exclude papers that result from your search terms. Systematic reviews require a flow diagram to show how this literature was included or excluded. This means that you have to be systematic in how you record weather a paper is included or excluded.

I'm my opinion, a systematic review is less subjective than a review where the author is essentially cherry picking the literature in order to tell the story as they see it. It also allows you to conduct statistical tests in order to answer some of your questions (a meta-analysis). Lastly, it avoids one of the oldest problems in science: confirmation bias: Simply looking for articles that back up your own view. Too many old reviews are really just written by academics seeking to confirm their standpoints, and so moving forwards we must place the emphasis on being objective through systematic reviews. This view is now growing so that increasing numbers of journals will only accept a review if it is systematic.

## Using Google Scholar
Google scholar has distinct advantages over other literature databases as it searches words inside the contents of an article, not just in the title, abstract and keywords (like Scopus and WoS). However, the output style of Google Scholar is infuriating as it resists being able to capture search results into a simple table (database) format. In addition, it does not support the boolean search operators (see part 2), and has a far broader range of literature that is included.

Help is at hand for those of you who would like to use Google Scholar and get a database output with the publication of [Publish or Perish software](https://harzing.com/resources/publish-or-perish) (Harzing 2007). The original purpose of this software was to calculate personal research metrics for promotion purposes, and it still does all this. But it also provides a single platform for searching many different literature databases: Scopus, Web of Science, Microsoft Academic, Google Scholar.

Google Scholar has an important attribute over many other databases in that it returns results in languages other than English. Using only English to provide the basis for your review might lead you to draw an unreasonable conclusion with a literature bias (see Nuñez & Amano 2021). Language is only one of several potential biases in a literature review (see Amano & Sutherland 2013).

@@

### Combining topics by combining people
The literature review that you conduct for your PhD will likely only have you doing the bulk of the work. But reviews become especially insightful when you combine topics together. To do this you may want to find somebody else who is an expert on the literature in that field. This might be somebody in your lab group. It might be somebody who is also just starting a PhD and is writing a literature review.

@@

## Don't try to do too much
Whatever your aims were before you started compiling the literature, remember to remain flexible as you proceed. it's hard to know exactly how many papers you will encounter on a particular subject unless you already know it very well. If you're prepared to remain flexible while conducting your review it may save you from overreaching. Always be open to reducing the scope of your review or seeing particular questions or issues while you're reading that may be better than your initial idea.

## In summary
if you come to the end of your PhD and have not yet done a literature review then now is the time. You should already know a large chunk of the literature in your particular area and you will be looking for future questions to tackle in a postdoc. Conducting a literature review at this stage in your career will not only help you to highlight unanswered questions and interesting topics, but it will get your name out there. Once you've done it you'll be glad you have.

>>>>

The first chapter of many PhDs is often a place for a literature review. Writing a review of the literature is actually a very useful task to undertake during your PhD, preferably towards the beginning. Firstly, it will get you familiar with all of the relevant literature in your field. It will give you some idea of the historical context of your subject area. And it's especially good to help you spot gaps in the literature that may not have been realised before.

@@

## Having a purpose
Writing a literature review is easiest when you have a clear purpose. The objective can be particularly tight, or can aim to use the literature in order to answer a broader question. Either way, having a clear purpose for your literature review will help you to define exactly what literature you're looking for.

Without an aim a literature review can be extremely daunting. There will be literally no end to the literature that you could include and it may well overwhelm you. If your question is more broad it will help to define a clear time period in which you are searching for literature. For example, you may want to just consider the last 10 years. Remember that the amount of literature is increasing almost exponentially and so if you are covering the same ground as a previous review a lot can happen in 10 years.

@@

## Systematic reviews
I prefer to conduct a systematic review. This is where you define your search criteria and the databases that you want to use and then include all of the literature that comes out. You will need to provide clear criteria for the inclusion of literature. As well as explaining exactly how you decide to include or exclude papers that result from your search terms. Systematic reviews require a flow diagram to show how this literature was included or excluded. This means that you have to be systematic in how you record whether a paper is included or excluded.

An important component of a systematic review is to include a Prisma flow-diagram [@moher2009preferred see Figure \@ref(fig:Prisma-flow)]. Many journals (especially those linked to PubMed) now insist that a Prisma flow-diagram has to be Figure 1 of any systematic review submitted. Simply replace the text with relevant numbers as appropriate in your systematic literature review. You can also add boxes as necessary. Note that the need to produce the Prisma flow-diagram means that you should keep good records of the results (and date) each time you conduct a literature search. You should also retain the actual search term used.

(ref:Prisma-flow) **The Prisma flow diagram after Moher et al (2009)**. Text inside the boxes is replaced by the author with the corresponding numbers of records, articles and studies as they were filtered during their systematic literature review, and (if appropriate) the final meta-analysis.

```{r Prisma-flow, echo=FALSE, fig.cap="(ref:Prisma-flow)"}
 knitr::include_graphics("figures/Prisma-flow.png")
```

In my opinion, a systematic review is less subjective than a review where the author is essentially cherry picking the literature in order to tell the story as they see it. It also allows you to conduct statistical tests in order to answer some of your questions (a meta-analysis). Lastly, it avoids one of the oldest problems in science: confirmation bias: Simply looking for articles that back up your own view. Too many old reviews are really just written by academics seeking to confirm their standpoints, and so moving forwards we must place the emphasis on being objective through systematic reviews. This view is now growing so that increasing numbers of journals will only accept a review if it is systematic.

## Using Google Scholar
Google scholar has distinct advantages over other literature databases as it searches words inside the contents of an article, not just in the title, abstract and keywords (like Scopus and WoS). However, the output style of Google Scholar is infuriating as it resists being able to capture search results into a simple table (database) format. In addition, it does not support the boolean search operators [[see](http://howtowriteaphd.com/databases.html) @measey2021how], and includes a far broader range of literature, and linking citations [@martin-martin2021google].

Help is at hand for those of you who would like to use Google Scholar and get a database output with the publication of [Publish or Perish software](https://harzing.com/resources/publish-or-perish) [@harzing2007publish]. The original purpose of this software was to calculate personal research metrics for promotion purposes, and it still does all this. But it also provides a single platform for searching many different literature databases: Scopus, Web of Science, Microsoft Academic (now discontinued), Google Scholar.

Google Scholar has an important attribute over many other databases in that it returns results in languages other than English. Using only English to provide the basis for your review might lead you to draw an unreasonable conclusion with a literature bias [@nunez2021monolingual]. Language is only one of several potential biases in a literature review [@amano2013four].

@@

### Combining topics by combining people
Reviews become especially insightful when you combine topics together. To do this you may want to find somebody else who is an expert on the literature in that field. This might be somebody in your lab group, or it might be someone in your wider [network](#networks). Bringing together a set of like-minded people for a review is a very powerful way in which to create an important and potentially powerful [network of collaborators](#networks).

@@

## Don't try to do too much
Whatever your aims were before you started compiling the literature, remember to remain flexible as you proceed. It's hard to know exactly how many papers you will encounter on a particular subject unless you already know it very well. If you're prepared to remain flexible while conducting your review it may save you from overreaching. Always be open to reducing the scope of your review or seeing particular questions or issues while you're reading that may be better than your initial idea.

## In summary
If you have got to this stage of your career and have not yet done a literature review then now is the time. You should already know a large chunk of the literature in your particular area and you will be looking for future questions to tackle in your early career. Conducting a literature review at this stage  will not only help you to highlight unanswered questions and interesting topics, but it will get your name out there. Once you've done it you'll be glad you did.
